---
title: "The DMV State Police are Racially Profiling People"
author: "Santiago Taborga"
institute: "st6374a@american.edu"
date: "04/09/2025"
output:
  xaringan::moon_reader:
    lib_dir: libs
    css: [default, fc, fc-fonts]
    nature:
      highlightStyle: github
      highlightLines: true
      ratio: "16:9"
---


```{r setup, include=FALSE}
## Libraries

# Setup -----

## Libraries
library(tidyverse)
library(stargazer)
library(modelsummary)
library(gt)
library(knitr)
library(kableExtra)

## Global options
knitr::opts_chunk$set(
  echo = FALSE, 
  prompt = FALSE,
  tidy = FALSE, 
  comment = NA,
  message = FALSE, 
  warning = FALSE,
  fig.align = 'center'
)

## Loading the data
md <- readRDS("yg821jf8611_md_statewide_2020_04_01.rds") %>%
  mutate(
    state = "Maryland"
  )

va <- readRDS("yg821jf8611_va_statewide_2020_04_01.rds") %>%
  mutate(
    state = "Virginia"
  )

## combining and cleanup
dmv <- full_join(md, va) %>%
  relocate(state, .after = date) %>%
  filter(
    !is.na(subject_race),
    !is.na(date),
    subject_race != "unknown"
  ) %>%
  mutate( 
    search_conducted = if_else(search_conducted == "TRUE", 1, 0),
    subject_race = factor(subject_race),
    year = lubridate::year(date)
  )

## graph settings
theme_classic2 = theme_classic() +
  theme(
    text = element_text(family = "Times"),
    plot.title = element_text(size = 9, face = "bold"),
    axis.title.x = element_text(size = 8),
    axis.text.x = element_text(size = 7),
    axis.text.y = element_text(size = 7),
    axis.title.y = element_text(size = 8),
    legend.text = element_text(size = 7),
    legend.title = element_text(size = 7, face = "bold"),
    legend.position = "bottom",
    legend.direction = "horizontal"
  )

custom_colors <- c(
  "asian/pacific islander" = "#488f31", 
  "black" = "#a5c796", 
  "hispanic" = "#f3a3a4",
  "white" = "#ffcc90",
  "other" = "#de425b"
)

custom_colors2 <- c(
  "2006" = "#AEC6CF",
  "2007" = "#FFB347",
  "2008" = "#B39EB5",
  "2009" = "#77DD77",
  "2010" = "#FF6961",
  "2011" = "#FDFD96",
  "2012" = "#CFCFC4",
  "2013" = "#F49AC2",
  "2014" = "#B0E0E6",
  "2015" = "#E6E6FA",
  "2016" = "#D9EAD3"
)
```

# Overview

  - **Motivation**

Studying racial profiling in the DMV area has the potential to reveal important insights into the intersections of race, law enforcement, and public policy. It can contribute to creating fairer policing practices, increase accountability, and ultimately help create a more equitable society for everyone, especially historically marginalized racial groups. The DMV area, with its political significance, demographic diversity, and proximity to national institutions, makes it an ideal place to study these critical issues.


  - **Central Findings**
  
    1. Racial profiling is prevalent when comparing white to non-white groups.
    
    2. When comparing specific racial sub-groups, racial profiling is still present, except for people that fall within the "Other" category.
    
    3. Group-specific racial profiling does not remain consistent across years.
    
    4. Hispanics are the most racially profiled gruop.

---
# Data

  - **Where it comes from**
  
    - The data was retrieved from the Stanford Open Policing Project.
    
    - The Project collects traffic stop information provided by state and local police departments in 42 states, wherever it was made available.
    
    -  Each jurisdiction provided stop data in idiosyncratic formats with varying levels of specificity due to the reporting requirements of each police department.
    
    
  - **What was done with it**
  
    - Virginia and Maryland's State Patrol
    
    - 8,505,026 observations
    
  - **Summary** 
  
    - *Unit of Analysis:* traffic stops
    
    - *Independent Variable:* Subject's Race
    
    - *Dependent Variable:* Search Conducted
    
    - *Time Period:* 2006-2016
    
    
---
# Hispanics Are Searched More Frequently

```{r, fig.width = 5, fig.height = 2.2, dpi = 300}
# profiling graph -----
profiling <- dmv %>%
  group_by(subject_race) %>%
  summarize (
    search_conducted = mean(search_conducted, na.rm = TRUE)
    ) %>%
  arrange(desc(search_conducted)) %>%
  mutate(subject_race = factor(subject_race, levels = subject_race)) %>%
  ggplot(aes(x = search_conducted, y = subject_race, fill = subject_race)) + 
  geom_bar(stat = "identity", color = "black") +
  labs(
    title = "% Search Conducted by Race",
    x = "Mean Search Conducted",
    y = ""
    ) +
  scale_x_continuous(
    labels = scales::percent_format(scale = 100),
    breaks = seq(0, 0.08, by = 0.02),
    limits = c(0, 0.08)) +
  theme_classic2 +
  theme(legend.title = element_blank(), 
        legend.text = element_blank(),
        axis.line.y = element_blank()) +
  scale_fill_manual(values = custom_colors,
                    guide = "none") +
  scale_color_manual(values = custom_colors)
profiling
```

---
# What Does A Comparison By Year Tell Us?

``` {r, fig.width = 5, fig.height = 2.2, dpi = 300}
# across years graph-----
profiling_year <- dmv %>%
  group_by(subject_race, year) %>%
  summarize(
    search_conducted = mean(search_conducted, na.rm = TRUE)
  ) %>%
  ggplot(aes(x = subject_race, y = search_conducted, fill = factor(year))) +
  geom_bar(stat = "identity", position = "dodge", color = "black") +
  labs(
    title = "% Search Conducted by Race and Year",
    x = "",
    y = "Mean Search Conducted",
    fill = "Year"
  ) +
  scale_y_continuous(
    labels = scales::percent_format(scale = 100),
    breaks = seq(0, 0.08, by = 0.02),
    limits = c(0, 0.08)) +
  theme_classic2 +
  scale_fill_manual(values = custom_colors2) +
  theme(
    legend.position = "right",
    legend.direction = "vertical",
    legend.box = "vertical",
    legend.key.size = unit(0.25, "cm"))

profiling_year
```

---
# Highlighted Rows = Asians/Pacific Islanders Being Searched the Most in That Year

``` {r}
# across years table-----
profiling_year_data <- dmv %>%
  group_by(subject_race, year) %>%
  summarize(
    search_conducted = mean(search_conducted, na.rm = TRUE) * 100
  ) %>%
  group_by(year) %>%
  mutate(max_value = max(search_conducted)) %>%
  ungroup()

## Create the table
profiling_year <- profiling_year_data %>%
  filter(subject_race == "asian/pacific islander") %>%
  gt() %>%
  cols_hide(columns = c(max_value, subject_race)) %>%
  tab_header(
    title = "% Search Conducted by Race and Year: Asian/Pacific Islander"
  ) %>%
  cols_label(
    subject_race = "Race",
    year = "Year",
    search_conducted = "Mean Search Conducted"
  ) %>%
  tab_style(
    style = cell_text(size = 8, weight = "bold"),
    locations = cells_title()
  ) %>%
    tab_style(
    style = cell_text(size = 8),
    locations = cells_column_labels()
  ) %>%
  tab_style(
    style = list(
      cell_text(size = 7, color = "black")
    ),
    locations = cells_body()
  ) %>%
  tab_style(
    style = list(
      cell_fill(color = "yellow"),
      cell_text(weight = "bold", color = "black")
    ),
    locations = cells_body(
      rows = search_conducted == max_value
    )
  ) %>%
  fmt_number(
    columns = c(search_conducted),
    decimals = 3,
    use_seps = TRUE
  )

profiling_year
```
---
# Highlighted Rows = Blacks Being Searched the Most in That Year

``` {r}
## Create the table
profiling_year <- profiling_year_data %>%
  filter(subject_race == "black") %>%
  gt() %>%
  cols_hide(columns = c(max_value, subject_race)) %>%
  tab_header(
    title = "% Search Conducted by Race and Year: Black"
  ) %>%
  cols_label(
    subject_race = "Race",
    year = "Year",
    search_conducted = "Mean Search Conducted"
  ) %>%
  tab_style(
    style = cell_text(size = 8, weight = "bold"),
    locations = cells_title()
  ) %>%
    tab_style(
    style = cell_text(size = 8),
    locations = cells_column_labels()
  ) %>%
  tab_style(
    style = list(
      cell_text(size = 7, color = "black")
    ),
    locations = cells_body()
  ) %>%
  tab_style(
    style = list(
      cell_fill(color = "yellow"),
      cell_text(weight = "bold", color = "black")
    ),
    locations = cells_body(
      rows = search_conducted == max_value
    )
  ) %>%
  fmt_number(
    columns = c(search_conducted),
    decimals = 3,
    use_seps = TRUE
  )

profiling_year
```
---
# Highlighted Rows = Hispanics Being Searched the Most in That Year

``` {r}
## Create the table
profiling_year <- profiling_year_data %>%
  filter(subject_race == "hispanic") %>%
  gt() %>%
  cols_hide(columns = c(max_value, subject_race)) %>%
  tab_header(
    title = "% Search Conducted by Race and Year: Hispanic"
  ) %>%
  cols_label(
    subject_race = "Race",
    year = "Year",
    search_conducted = "Mean Search Conducted"
  ) %>%
  tab_style(
    style = cell_text(size = 8, weight = "bold"),
    locations = cells_title()
  ) %>%
    tab_style(
    style = cell_text(size = 8),
    locations = cells_column_labels()
  ) %>%
  tab_style(
    style = list(
      cell_text(size = 7, color = "black")
    ),
    locations = cells_body()
  ) %>%
  tab_style(
    style = list(
      cell_fill(color = "yellow"),
      cell_text(weight = "bold", color = "black")
    ),
    locations = cells_body(
      rows = search_conducted == max_value
    )
  ) %>%
  fmt_number(
    columns = c(search_conducted),
    decimals = 3,
    use_seps = TRUE
  )

profiling_year
```
---
# Highlighted Rows = Whites Being Searched the Most in That Year

``` {r}
## Create the table
profiling_year <- profiling_year_data %>%
  filter(subject_race == "white") %>%
  gt() %>%
  cols_hide(columns = c(max_value, subject_race)) %>%
  tab_header(
    title = "% Search Conducted by Race and Year: White"
  ) %>%
  cols_label(
    subject_race = "Race",
    year = "Year",
    search_conducted = "Mean Search Conducted"
  ) %>%
  tab_style(
    style = cell_text(size = 8, weight = "bold"),
    locations = cells_title()
  ) %>%
    tab_style(
    style = cell_text(size = 8),
    locations = cells_column_labels()
  ) %>%
  tab_style(
    style = list(
      cell_text(size = 7, color = "black")
    ),
    locations = cells_body()
  ) %>%
  tab_style(
    style = list(
      cell_fill(color = "yellow"),
      cell_text(weight = "bold", color = "black")
    ),
    locations = cells_body(
      rows = search_conducted == max_value
    )
  ) %>%
  fmt_number(
    columns = c(search_conducted),
    decimals = 3,
    use_seps = TRUE
  )

profiling_year
```
---
# Highlighted Rows = Others Being Searched the Most in That Year

``` {r}
## Create the table
profiling_year <- profiling_year_data %>%
  filter(subject_race == "other") %>%
  gt() %>%
  cols_hide(columns = c(max_value, subject_race)) %>%
  tab_header(
    title = "% Search Conducted by Race and Year: Other"
  ) %>%
  cols_label(
    subject_race = "Race",
    year = "Year",
    search_conducted = "Mean Search Conducted"
  ) %>%
  tab_style(
    style = cell_text(size = 8, weight = "bold"),
    locations = cells_title()
  ) %>%
    tab_style(
    style = cell_text(size = 8),
    locations = cells_column_labels()
  ) %>%
  tab_style(
    style = list(
      cell_text(size = 7, color = "black")
    ),
    locations = cells_body()
  ) %>%
  tab_style(
    style = list(
      cell_fill(color = "yellow"),
      cell_text(weight = "bold", color = "black")
    ),
    locations = cells_body(
      rows = search_conducted == max_value
    )
  ) %>%
  fmt_number(
    columns = c(search_conducted),
    decimals = 3,
    use_seps = TRUE
  )

profiling_year
```
