---
title: "The DMV State Police are Racially Profiling People"
author: "Santiago Taborga"
institute: "st6374a@american.edu"
output:
  beamer_presentation:
    theme: "Malmoe"
    colortheme: "beaver"
fontsize: 11pt
classoption: "aspectratio=169"
---

```{r setup, include=FALSE}
## Libraries

# Setup -----

## Libraries
library(tidyverse)
library(stargazer)
library(modelsummary)
library(gt)
library(knitr)
library(kableExtra)

## Global options
knitr::opts_chunk$set(
  echo = FALSE, 
  prompt = FALSE,
  tidy = FALSE, 
  comment = NA,
  message = FALSE, 
  warning = FALSE,
  fig.align = 'center'
)

## Loading the data
md <- readRDS("yg821jf8611_md_statewide_2020_04_01.rds") %>%
  mutate(
    state = "Maryland"
  )

va <- readRDS("yg821jf8611_va_statewide_2020_04_01.rds") %>%
  mutate(
    state = "Virginia"
  )

## combining and cleanup
dmv <- full_join(md, va) %>%
  relocate(state, .after = date) %>%
  filter(
    !is.na(subject_race),
    !is.na(date),
    subject_race != "unknown"
  ) %>%
  mutate( 
    search_conducted = if_else(search_conducted == "TRUE", 1, 0),
    subject_race = factor(subject_race),
    year = lubridate::year(date)
  )

## graph settings
theme_classic2 = theme_classic() +
  theme(
    text = element_text(family = "Times"),
    plot.title = element_text(size = 10, face = "bold"),
    axis.title.x = element_text(size = 9),
    axis.text.x = element_text(size = 8),
    axis.text.y = element_text(size = 8),
    axis.title.y = element_text(size = 9),
    legend.text = element_text(size = 8),
    legend.title = element_text(size = 8, face = "bold"),
    legend.position = "bottom",
    legend.direction = "horizontal"
  )

custom_colors <- c(
  "asian/pacific islander" = "#488f31", 
  "black" = "#a5c796", 
  "hispanic" = "#f3a3a4",
  "white" = "#ffcc90",
  "other" = "#de425b"
)
```

# Overview

## Overview

  - Motivation

Studying racial profiling in the DMV area has the potential to reveal important insights into the intersections of race, law enforcement, and public policy. It can contribute to creating fairer policing practices, increase accountability, and ultimately help create a more equitable society for everyone, especially historically marginalized racial groups. The DMV area, with its political significance, demographic diversity, and proximity to national institutions, makes it an ideal place to study these critical issues.

  - Central Findings
    1. Racial profiling is prevalent when comparing white to non-white groups.
    2. When comparing specific racial sub-groups, racial profiling is still present, except for people that fall within the "Other" category.
    3. Group-specific racial profiling does not remain consistent across years.
    4. Hispanics are the most racially profiled gruop.

# Data

## Data

  - Where it comes from
    - The data was retrieved from the Stanford Open Policing Project.
    - The Project collects traffic stop information provided by state and local police departments in 42 states, wherever it was made available.
  - What was done with it
    - Virginia and Maryland's State Patrol
    - 8,505,026 observations
    

# Analysis

## Analysis Pt. 1

```{r, fig.dim=c(5,2.2)}
# profiling graph -----
profiling <- dmv %>%
  group_by(subject_race) %>%
  summarize (
    search_conducted = mean(search_conducted, na.rm = TRUE)
    ) %>%
  ggplot(aes(x = search_conducted, y = subject_race, fill = subject_race)) + 
  geom_bar(stat = "identity", color = "black") +
  labs(
    title = "% Search Conducted by Race",
    x = "Mean Search Conducted",
    y = ""
    ) +
  scale_x_continuous(
    labels = scales::percent_format(scale = 100),
    breaks = seq(0, 0.08, by = 0.02),
    limits = c(0, 0.08)) +
  theme_classic2 +
  theme(legend.title = element_blank(), 
        legend.text = element_blank(),
        axis.line.y = element_blank()) +
  scale_fill_manual(values = custom_colors,
                    guide = "none") +
  scale_color_manual(values = custom_colors)
profiling
```

## Analysis Pt. 1

```{r}
# regression models -----

## making data race-specific
overall <- dmv %>%
  mutate(
    subject_race = if_else(subject_race == "white", "white", "non-white")
  )

asian <- dmv %>%
  filter(subject_race == c("asian/pacific islander", "white"))

black <- dmv %>%
  filter(subject_race == c("black", "white"))

hispanic <- dmv %>%
  filter(subject_race == c("hispanic", "white"))

other <- dmv %>%
  filter(subject_race == c("other", "white"))

## estimates
mods <- 
  list(
    "Overall" = lm(search_conducted ~ subject_race, overall),
    "Asian/pacific islander" = lm(search_conducted ~ subject_race, asian),
    "Black" = lm(search_conducted ~ subject_race, black),
    "Hispanic" = lm(search_conducted ~ subject_race, hispanic),
    "Other" = lm(search_conducted ~ subject_race, other)
  )

## displaying all models
table <- modelsummary(
  mods,
  coef_rename = c(
    "subject_racewhite" = "White",
    "subject_raceother" = "Other"
  ),
  gof_map = c('nobs'),
  output = 'gt',
  star = TRUE
)

## Add header above the table
table %>%
  tab_spanner(
    label = "Race",
    columns = c(`Overall`, `Asian/pacific islander`, `Black`, `Hispanic`, `Other`)
  )
```

## Analysis Pt. 2

``` {r, fig.dim=c(5,2.2)}
# across years graph-----
profiling_year <- dmv %>%
  group_by(subject_race, year) %>%
  summarize(
    search_conducted = mean(search_conducted, na.rm = TRUE)
  ) %>%
  ggplot(aes(x = factor(year), y = search_conducted, fill = subject_race)) +
  geom_bar(stat = "identity", position = "dodge", color = "black") +
  labs(
    title = "% Search Conducted by Race and Year",
    x = "",
    y = "Mean Search Conducted",
    fill = "Race"
  ) +
  scale_y_continuous(
    labels = scales::percent_format(scale = 100),
    breaks = seq(0, 0.08, by = 0.02),
    limits = c(0, 0.08)) +
  theme_classic2 +
  scale_fill_manual(values = custom_colors) +
  theme(legend.key.size = unit(0.5, "cm"))

profiling_year
```

## Analysis Pt. 2
``` {r}
# across years table-----
profiling_year_data <- dmv %>%
  group_by(subject_race, year) %>%
  summarize(
    search_conducted = mean(search_conducted, na.rm = TRUE) * 100
  ) %>%
  group_by(year) %>%
  mutate(max_value = max(search_conducted)) %>%
  ungroup()

## Create the table
profiling_year <- profiling_year_data %>%
  gt() %>%
  cols_hide(columns = c(max_value)) %>%
  tab_header(
    title = "% Search Conducted by Race and Year"
  ) %>%
  cols_label(
    subject_race = "Race",
    year = "Year",
    search_conducted = "Mean Search Conducted"
  ) %>%
  tab_style(
    style = cell_text(size = 8, weight = "bold"),
    locations = cells_title()
  ) %>%
    tab_style(
    style = cell_text(size = 8),
    locations = cells_column_labels()
  ) %>%
  tab_style(
    style = list(
      cell_text(size = 7, color = "black")
    ),
    locations = cells_body()
  ) %>%
  tab_style(
    style = list(
      cell_fill(color = "yellow"),
      cell_text(weight = "bold", color = "black")
    ),
    locations = cells_body(
      rows = search_conducted == max_value
    )
  ) %>%
  fmt_number(
    columns = c(search_conducted),
    decimals = 3,
    use_seps = TRUE
  )

profiling_year
```